---
title: "Red Wine Analysis"
author: "Beth Morrissette"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#Load libraries needed for analysis

library(ggplot2)
library(tidyverse)
library(scales)
library(gridExtra)
library(GGally)
library(RColorBrewer)
library(corrplot)
library(Hmisc)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data

wine=read.csv("wineQualityReds.csv")
```

> **Introduction**

This dataset is related to red variants of the Portuguese "Vinho Verde" wine.
There are 13 variables in this data set including acidity measures, density,
and sulphates.


# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
summary(wine)

```
>
>

```{r echo=FALSE}
str(wine)
```

Added a Quality Factor variable to better reflect what it is measuring.
Created a category variable, "Wine Rating" to bucket the scores.  1-3 = "Poor"
4-6 = "Average", 7-9 = Good, 10 = Excellent

```{r, echo=FALSE}
wine$rating = case_when(
  wine$quality < 4 ~ "Poor",
  wine$quality < 7 ~ "Average",
  wine$quality < 9 ~ "Good",
  wine$quality == 10 ~ "Excellent"
)

wine$quality.factor = factor(wine$quality, ordered = TRUE)
wine$rating = factor(wine$rating,levels = c('Poor', 'Average', 'Good','Excellent', ordered = TRUE))
```
>
>

```{r, echo=FALSE}
str(wine)
```


```{r, echo = FALSE, warning=FALSE}
ggplot(data = wine, aes(x=rating)) +
  geom_bar()+
  xlab('Rating') +
  ggtitle('Wine Rating Category')
```


Majority of wines fall in to the "Average" (4-6) Rating category.  Took a look
at the rating values to see breakdown


```{r, echo = FALSE}
ggplot(data=wine,aes(x=quality)) +
  geom_bar()+
  scale_x_continuous(breaks = seq(0,8,1))+
  xlab('Quality Score') +
  ggtitle('Quality Score Histogram')
```


```{r echo=FALSE, warning=FALSE}
alcohol1 = ggplot(data = wine,aes(x=alcohol)) +
  geom_histogram(binwidth = 0.25, color='blue') +
  xlab('% Alcohol Content')+
  ggtitle("%Alcohol Histogram") 

alcohol2=ggplot(data = wine,aes(x=alcohol)) +
  geom_histogram(binwidth = 0.25, color='blue') +
  scale_x_continuous(limits = c(7.5,12))+
  xlab('% Alcohol Content')+
  ggtitle("%Alcohol Histogram (Axis Limits)") 

grid.arrange(alcohol1, alcohol2, ncol=2)
```


Percentage of alcohol content is skewed to the left. But this smooths out
somewhat with outliers removed


```{r, echo=FALSE, warning=FALSE}
grid1=ggplot(data = wine, aes(x = sulphates )) +
  geom_histogram(binwidth = 0.05) +
  xlab('Potassium Sulphate - g / dm3)') +
  ggtitle('Sulphates Histogram')
grid2=ggplot(data = wine, aes(x = sulphates )) +
  geom_histogram(binwidth = 0.05) +
  scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x)) +
   xlab('Log10 Scale Potassium Sulphate - g / dm3)') +
  ggtitle('Sulphates Histogram', subtitle = 'Log10')

grid.arrange(grid1,grid2,ncol=2)
```

Long tailed - changed axis to log10 and removed outliers (>1.5) and smooths to
a more normal distribution

```{r echo=FALSE, warning=FALSE}
ph1 = ggplot(data = wine, aes(x=pH)) +
  geom_histogram(binwidth = 0.05, fill = 'gray', color = 'black') +
  ggtitle('Wine Acidity', subtitle = 'pH') +
  geom_vline(xintercept = mean(wine$pH), colour="blue", linetype = "longdash")
  
ph2 = ggplot(data = wine, aes(x=pH)) +
  geom_histogram(binwidth = 0.05, fill = 'gray', color = 'black') +
  scale_x_continuous(limits = c(3, 3.7))+
  ggtitle('Wine Acidity (Outliers Removed)', subtitle = 'pH') +
  geom_vline(xintercept = mean(wine$pH), colour="blue", linetype = "longdash")

grid.arrange(ph1,ph2, ncol=2)
```


Mean line added - almost a perfect normal distribution.  It could be folded in half on the mean

```{r, echo=FALSE, warning=FALSE}
density1 = ggplot(data = wine, aes(x = density)) +
  geom_histogram(binwidth = 0.001) +
  xlab('Density (g/cm^3)') +
  ggtitle('Wine Density Histogram')

density2 = ggplot(data = wine, aes(x=density)) +
  geom_histogram(binwidth = 0.001) + 
  scale_x_continuous(limits = c(0.9901, 1.004))+
   xlab('Density (g/cm^3)') +
  ggtitle('Wine Density Histogram', subtitle = 'Outliers Removed')

grid.arrange(density1, density2, ncol = 2)
```


```{r, echo=FALSE, warning=FALSE}
total.sulfur1 = ggplot(data = wine, aes(x = total.sulfur.dioxide)) +
  geom_histogram(binwidth = 10) +
  ggtitle('Total Sulfur Dioxide')

total.sulfur2 = ggplot(data = wine, aes(x = total.sulfur.dioxide)) +
  geom_histogram(binwidth = 0.05) +
  scale_x_log10(limits = c(5,200)) +
  ggtitle('Total Sulfur Dioxide', subtitle = 'Log10 X-Axis')

grid.arrange(total.sulfur1,total.sulfur2, ncol = 2)

```

After removing outlier (~300) and changing to a log 10 scale, we see a varied 
but more even distribution


```{r, echo=FALSE, warning=FALSE}
free.sulfur1 = ggplot(data = wine, aes(x=free.sulfur.dioxide)) +
  geom_histogram(binwidth = 2) + 
  ggtitle('Free Sulfur Dioxide')

free.sulfur2 = ggplot(data = wine, aes(x=free.sulfur.dioxide)) +
  geom_histogram(binwidth = 0.08) + 
   scale_x_log10(limits = c(2,90))+
  ggtitle('Free Sulfur Dioxide', subtitle = 'Log10 Axis')
  
  grid.arrange(free.sulfur1, free.sulfur2, ncol = 2)
```

Free Sulfur Dioxide looks similar to the total sulfur dioxide which makes sense
since it's a part of the total.


```{r echo=FALSE, warning=FALSE}

fixed.acid1 = ggplot(data = wine,aes(x=fixed.acidity)) +
  geom_histogram(binwidth = 1, color = 'black') +
  xlab('Fixed Acidity (Tartaric Acid - g / dm^3)') +
  ggtitle('Fixed Acidity Histogram')

fixed.acid2 = ggplot(data = wine,aes(x=fixed.acidity)) +
  geom_histogram(binwidth = 1, color = 'black') +
  scale_x_continuous(lim = c(4,14)) +
  xlab('Fixed Acidity (Tartaric Acid - g / dm^3)') +
  ggtitle('Fixed Acidity Histogram', subtitle = "Axis Limits")

grid.arrange(fixed.acid1, fixed.acid2, ncol = 2)
  
```

Fixed acidity had some outlier like data on the right so set limits to better see data


```{r echo = FALSE, warning=FALSE}

 fixed.acid1 = ggplot(data = wine, aes(x=volatile.acidity)) +
  geom_histogram(binwidth = 0.05, color = 'red') +
  xlab('Volatile Acidity (Acetic Acid - g / dm^3)') +
  ggtitle('Volatile Acidity')


fixed.acid2 = ggplot(data = wine, aes(x=volatile.acidity)) +
  geom_histogram(binwidth = 0.05, color = 'red') +
  xlab('Volatile Acidity (Acetic Acid - g / dm^3)') +
  xlim(0,quantile(wine$volatile.acidity,0.99)) +
  ggtitle('Volatile Acidity', subtitle = 'Top 99%')

grid.arrange(fixed.acid1, fixed.acid2, ncol=2)

```

Again, with volatile acidity, there were some outliers really pulling the curve,
so removed the top 1% from the plot


```{r echo = FALSE, warning=FALSE}

citric1 = ggplot(data = wine, aes(x=citric.acid)) +
  geom_histogram(binwidth = 0.05, color = 'white') +
  ggtitle('Citric Acid') +
  xlab('Citric Acid (g / dm^3)')

citric2 = ggplot(data = wine, aes(x=citric.acid)) +
  geom_histogram(mapping = aes(y = stat(density)), alpha=0.5, colour='black',
                 binwidth =0.05 ) +
  xlim(0,0.8) +
  stat_function(fun = dnorm, color = 'red', args = list(mean(wine$citric.acid), sd(wine$citric.acid))) +
  xlab("Citric Acid")+
  ggtitle("Citric Acid Density")
  

grid.arrange(citric1, citric2, ncol = 2)

```


With the top 1% removed, the Citric Acid looks almost bimodal


```{r echo = FALSE, warning=FALSE}

sugar1 = ggplot(data = wine, aes(x=residual.sugar)) +
  geom_histogram(binwidth = 0.5, fill = 'blue', color='gray') +
  xlab('Residual Sugar (g / dm^3)') +
  ggtitle('Residual Sugar Histogram')

sugar2 = ggplot(data = wine, aes(x=residual.sugar)) +
  geom_histogram(binwidth = 0.5, fill = 'blue', color='gray') +
  scale_x_continuous(limits = c(1,7)) +
  xlab('Residual Sugar (g / dm^3)') +
  ggtitle('Residual Sugar Histogram', subtitle = 'Outliers Removed') 

grid.arrange(sugar1, sugar2, ncol = 2)
```

Even with outliers removed, the sugar histogram is still skewed left.

```{r, echo=FALSE, warning=FALSE}
chloride1 = ggplot(data = wine, aes(x=chlorides)) + 
  geom_histogram(binwidth = 0.001)  +
  ggtitle('Chlorides')

chloride2 = ggplot(data = wine, aes(x=chlorides)) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limits = c(0.12, 0.2)) +
  ggtitle('Chlorides', subtitle = 'Outliers Removed')

grid.arrange(chloride1, chloride2, ncol = 2)
  
```

#Majority of the data falls between 0.07 and 0.09 so fairly concentrated with
#outliers

# Univariate Analysis

### What is the structure of your dataset?

This dataset is fairly well distributed in general although several variables
have outliers easily seen in the various plots.  

### What is/are the main feature(s) of interest in your dataset?
The biggest question is what factors in wine, affect quality?  Quality in 
this data set is a subjective rating applied to the wine (likely based on taste)

Based on this, I would like to take a closer look at the relationship between
Potassium Sulfate levels and Wine Quality/Rating.  There is much debate both
on the flavor as well as the health frontier about this preservative so it will 
be interesting to see if there is, indeed, a relationship.

Residual sugar -- does a sweeter wine equal a higher quality wine or vice versa?

What effect does acidity (pH) have on the quality of wine?

Finally, would like to see the effect of alcohol content on the quality of the
wine.


### What other features in the dataset do you think will help support your investigation? \
Transforming the quality item to an ordered factor and adding the bucket will 
help more easily compare them to other factors.

### Did you create any new variables from existing variables in the dataset?

I created a new "Rating" variable - an ordered factor bucketing the different
quality ratings.

### Of the features you investigated, were there any unusual distributions? \
Several of the plots had outliers that seemed to be artificially pulling the 
data to the left.  I used several transformations to address this.  Some were
simply a matter of adding limits to the x-axis - both directly using the xlim
parameter and indirectly by using the quantile function to remove the top 1%.  
In some cases, a logarithmic scale for the x-axis was helpful in better seeing
the distribution of the data.


# Bivariate Plots Section

```{r, echo=FALSE, warning=FALSE}
ggplot(data = wine, aes(x=rating, y=sulphates)) +
  geom_boxplot(fill = "cornflowerblue", 
               alpha = 0.5,outlier.colour="blue", outlier.shape=7) +
  ylim(0,1.5) +
  ylab("Sulphates (g/dm3)") +
  xlab('Wine Rating Category')+
  ggtitle('Sulphates by Wine Rating Category') +
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 2)


```

We see a higher level of sulfates as the category improves.  Still a lot of 
outliers in the average category


```{r, echo=FALSE,warning=FALSE}
ggplot(data = wine, aes(x=rating, y=residual.sugar))+
  geom_jitter(alpha = 1/15, color="blue") +
  scale_y_continuous(limits = c(0,12)) +
  xlab('Wine Rating') +
  ylab('Residual Sugar (gm / dm^3)') +
  ggtitle('Residual Sugar Across Ratings')
  
```

Doesn't appear to be much of a correlation here.  

```{r, echo=FALSE}
average_plot = wine%>%
  group_by(rating)%>%
  summarise(n=n(),
            mean= mean(pH),
            sd = sd(pH),
            se = sd / sqrt(n),
            ci = qt(0.975, df=n - 1, )* sd / sqrt(n))
```


```{r, echo=FALSE, warning=FALSE}
ggplot(data = average_plot, aes(x = rating, y = mean, group=1)) +
  geom_point(size = 3) +
  geom_line(linetype='longdash') +
  geom_errorbar(aes(ymin = mean - se, 
                    ymax = mean + se), 
                width = .1) 
```

Created additional dataframe with summarized pH data. Appears that 
a lower average pH (more acidic) tends to positively impact quality


```{r, echo=FALSE, warning=FALSE}

ggplot(data = wine,aes(x=quality.factor, y=alcohol)) + 
  geom_jitter(color = 'red', alpha = 1/20) +
  xlab("Quality Score") +
  ylab("Alcohol Content (% by Volume)") +
  scale_y_continuous(breaks = seq(0,15,1))+
  ggtitle("Alcohol Content by Quality Score")

```

Almost no alcohol content in the wines rated "poor'.  Higher concentration
at the 9-10 mark in the average wines and a somewhat more
diffuse distribution in the good.  Looking at the raw scores, there's less
alcohol content as the score goes up from 5


```{r, echo=FALSE}
ggplot(data = wine, aes(x=quality.factor, y = volatile.acidity)) + 
  geom_boxplot()
```

Volatile acidity is considered a sign of spoilage - your wine is turning to
vinegar at that point so it makes sens to see lower volatile acidity in higher 
quality ratings


```{r, echo=FALSE}
ggplot(data = wine, aes(x=rating, y=density)) +
  geom_boxplot() +
  xlab('Wine Rating')+
  ylab('Density(g/dm^3)') +
  ggtitle('Density over Quality')
```

The density of wine is primarily determined by the concentration of alcohol, 
sugar, glycerol, and other dissolved solids. Density in better wines is lower.

```{r, echo=FALSE}
ggplot(data = wine, aes(x=rating, y=chlorides)) +
  geom_jitter(color='orange', alpha = 1/20) +
  xlab("Wine Rating")+
  ylab("Sodium Chloride (g/dm^3)") +
  ggtitle("Sodium Chloride over Quality")
```

Nothing conclusive here.  Concentration of sodium chloride could be because
the majority of wines fall under the average category


```{r, echo=FALSE}
ggplot(data = wine, aes(x=residual.sugar, y=sulphates)) +
  geom_point(alpha = 1/5, position = position_jitter(h=0),
             color = "orange") +
  stat_summary(geom = 'line',fun='mean') +
  coord_cartesian(xlim = c(1,3), ylim = c(0.4,1.25)) +
  xlab("Residual Sugar") +
  ylab("Sulphates") +
  ggtitle("Sulphates Over Residual Sugar (with mean)")
```

No real relationship here - if there were I'd expect the points of the mean
to be inline with the concentration of sugar.


```{r, echo=FALSE, warning=FALSE}
ggplot(data = wine, aes(x=residual.sugar, y = alcohol)) +
  geom_point()+
 geom_smooth(method = "lm", formula = y ~ x, size = 1, se = TRUE,
    colour = "violet") +
  scale_x_continuous(limits = c(0,7), breaks = seq(0,10,1)) +
  xlab("Residual Sugar") +
  ylab("Alcohol Content(% by volume)") +
  ggtitle("Alcohol Content over Residual Sugar", subtitle = "With Conditional Means")
```

Doesn't appear to be a linear relationship between alcohol and sugar even 
though both are part of density and have similar relationships with quality

```{r, echo = FALSE}
ggplot(data = wine, aes(x=fixed.acidity, y = pH)) +
  geom_point(alpha = 0.07) + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1, se = TRUE,
    colour = "violet") +
  xlab("Fixed Acidity") +
  ylab("pH") + 
  ggtitle("pH over Fixed Acidity", subtitle = "With Conditional Means")
  
```


First a heat map to look at correlation amongst all the variables.  Wine Quality
(factor) and wine rating (catgory) were excluded due to non numeric status.

```{r echo=FALSE, Multivariate_Plots}
wine.subset=wine[ , !names(wine) %in% c("X", "quality.factor", "rating")]
res=cor(wine.subset)

col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(res, method="color", col=col(200),  
         type="upper", order="hclust", 
         addCoef.col = "black", # Add coefficient of correlation
         tl.col="black", tl.srt=45, #Text label color and rotation
         diag=FALSE #OMit matching with itself
         )
```



# Bivariate Analysis

Alcohol content increases as the quality of the wine improves.  Sulphates go up
as quality increases.  Interestingly, there didn't seem to be much of a 
relationship between residual sugar and quality.  

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?  
Yes, there is a relationship between fixed acidity and pH. Since
pH is a measure of acidity, this is expected. 
Wine density does appear to have an inverse relationship with quality (less 
dense = higher quality)

### What was the strongest relationship you found?

Visually, the strongest relationship seen was between pH and quality.  Further 
analysis will be needed to see if there is true correlation

# Multivariate Plots Section



```{r, echo=FALSE}
df.cor = data.frame(category = c("Perfect Negative", "Volatile Acidity", 
                             "Total Sulfur Dioxide", "Density",
                             "Chlorides",
                             "pH",
                             "Free Sulfer Dioxide",
                             "None",
                             "Residual Sugar",
                             "Fixed Acidity",
                             "Citric Acid",
                             "Sulphates",
                             "Alcohol",
                             "Perfect Positive"),
                    correlation = c(-1, -0.39, -0.19, -0.17, -0.13, -0.06,
                                    -0.05, 0, 0.1, 0.12, 0.23, 0.25, 0.48, 1))

df.cor$category = factor(df.cor$category, levels = df.cor$category)

```

```{r, echo=FALSE}

ggplot(data = df.cor, aes(x = category, y = correlation)) +
  geom_col(aes(group=category)) +
  scale_x_discrete(labels=df.cor$category) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  

```




```{r, echo=FALSE}
ggplot(data = wine,
       aes(y = density, x = alcohol,
           color = rating)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", formula = (y~x),se = FALSE,size=1)  +
   scale_color_brewer(type='div',
                   guide=guide_legend(title='Quality'), palette = 'Accent')
 
```

```{r, echo = FALSE}
ggplot(data = wine, aes(x=alcohol, y=sulphates, color = rating)) + 
  geom_point() +
  scale_color_brewer(type='div',
                   guide=guide_legend(title='Quality'), palette = 'Accent')
```


# Multivariate Analysis

The strongest correlation with quality comes from alcohol content - at 0.48 it
is a medium correlation.  This is followed by volatile acidity at -0.39.  

There were some interesting interactions with volatile acid and pH but they 
didn't appear to have much correlation with quality.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

ggplot(data = df.cor, aes(x = category, y = correlation, fill = category)) +
  geom_col(aes(group=category)) +
 # scale_x_discrete(labels=df.cor$category) + 
  scale_color_brewer(type = 'seq', palette = 'Dark2') +
  ggtitle('Correlation with Quality', subtitle = 'All Factors') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

### Description One

This plot shows the correlation of each variable with quality on a scale
from perfect negative to perfect positive.  It is easy to see here that the 2
variables with the strongest impact on quality is Alcohol Content and Volatile
Acidity.  


### Plot Two
```{r echo=FALSE, Plot_Two}


ggplot(data = wine, aes(x=alcohol, y = sulphates, color = rating)) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'qual', palette = 'Set1',
    guide = guide_legend(title = 'Rating', reverse = F,
    override.aes = list(alpha = 1, size = 2)))  +
  ylab('Sulphates (g/dm^3)') +
  xlab("Alcohol Content (% by volume)") +
  ggtitle("Effect of Sulphates and Alcohol Content on Quality")

```


### Description Two

This plot shows the relationship between alcohol content and sulphates and 
their joined impact on quality.  Better wines have a higher alcohol content and
a higher sulphate content.  


### Plot Three

```{r, echo = FALSE}
ggplot(data = average_plot, aes(x = rating, y = mean, group=1)) +
  geom_point(size = 3, color = 'blue') +
  geom_line(linetype='longdash') +
  geom_errorbar(aes(ymin = mean - se, 
                    ymax = mean + se), 
                width = .1) +
  xlab('Rating Score') +
  ylab("Mean pH") +
  ggtitle('Mean pH by Rating with 95% Confidence Interval')

```



### Description Three

This plot shows the clear relationship between acidity and wine quality. The
mean pH score goes down markedly from poor to average and then again, less
dramatically, from average to good.  There is a pretty large confidence interval
on the poor rating and a fairly large one on the good rating.  This may be 
reflective of the sample being heavily distributed in the average rating.

------

# Reflection

This was an interesting opportunity to explore a data set. One of the hurdles faced
with this data set is the distribution of quality and it's concentration in the
average category.  If data was more evenly distributed, we might have a more 
clear picture of the factors which impact quality.

Alcohol content definitely played the leading role in determining quality followed
by volatile acid.  It was surprising to me that sugar content played less of 
a role in determining quality.  I had assumed that a sweeter wine would be
better liked but this was not shown in the data.

The rating scale used is completely subjective and a single integer.  No 
information is given on how it is determined other than personal taste.  For
future, it would be nice to have a data set where quality was a score based
on several factors - possibly weighted.  A larger, more evenly distributed 
sample might also help make relationships and causality more obvious.


> **Materials Used**
>
> DataNovia - GGPlot Section - https://www.datanovia.com/en/blog/category/ggplot2/
> R Documentation (STAT Function - https://www.rdocumentation.org/packages/ggplot2/versions/0.9.0/topics/stat_function)
> UC Davis Waterhouse Lab (https://waterhouse.ucdavis.edu/whats-in-wine)